---
title: "PALL visualization data"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    theme: united
    code_folding: hide
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: "Gianluca Alessio Mariani"

params:
  manual_timestamp: "250929_0900"
  project: "pall"
  clinical_data: "/hpcnfs/scratch/P_DIMA_PALL/report_ALL/data/clinical_data.csv"
  id_data: "/hpcnfs/scratch/P_DIMA_PALL/report_ALL/data/PATIENTS_ID.csv"
  gene_table1: "/hpcnfs/scratch/P_DIMA_PALL/report_ALL/data/PALL_WES_PrimaryTumors/FINAL_DATASET.csv"
  gene_table2: "/hpcnfs/scratch/P_DIMA_PALL/report_ALL/data/PALL_WES_RelapseTumors/FINAL_DATASET.csv"

  refgene_file: "/hpcnfs/scratch/DIMA/gmariani/__softwares/gistic2/gistic2/refgenes/hg38.UCSC.add_mir.160920/gistic2.refgene.hg38.UCSC.add_miR.160920.mat"
  cytoband_file: "/hpcnfs/scratch/DIMA/gmariani/_databases/reference_genomes/hg38/cytoBand_hg38_merged.txt"
  
  data_dir: "/hpcnfs/scratch/P_DIMA_PALL/report_ALL/scripts_report/R_project/data"
  results_dir: "/hpcnfs/scratch/P_DIMA_PALL/report_ALL/results"
  
  maf_tumor_dir: "/hpcnfs/scratch/P_DIMA_PALL/wes/250912_rerun_dragen4.4.4/annotations/variant_annotations/250916/annotation/somatic"
  cnv_dir: "/hpcnfs/scratch/P_DIMA_PALL/report_ALL/results/250923_1647_pall/250923/annotation/cnv"
  raw_samples_dir: "/hpcnfs/scratch/P_DIMA_PALL/wes/250912_rerun_dragen4.4.4"

  cnvkit_out_dir: "/hpcnfs/scratch/P_DIMA_PALL/report_ALL/results/250918_1626_cnvkit_pall"
  
  run_gistic2: FALSE
  num_top_genes: 25
  num_top_regions: 40
  type2use4cn_onco: "anno" #either "anno" or "seg"
  thr4gain: 0.30
  thr4amp: 0.90 
  thr4loss: -0.30 
  thr4deep: -0.90 
---

# Maftools Report

This report has been generated with R.

It contains the analysis for the PALL project, performed on a small set of 21 patients.

For each patient a DNA sample was collected at times t = diagnosis and subsequent relapses.

WES sequencing, variant calling and annotation was performed in previous steps to obtain the data presented in this report.

In particular the tools used to generate the input data are:

  - DRAGEN (4.4.4): sequencing, variant calling and CNV annotation
  
  - Variantalker (beta): small variants annotation
  
  - CNVkit (0.9.12): additional CNV annotation

```{r setup_knit, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	fig.align = "center",
	fig.height = 8,
	fig.width = 10,
	message = FALSE,
	warning = FALSE,
	autodep = TRUE,
	cache = FALSE,
	dev = c("png", "pdf")
)
```

```{r load_libraries_and_setup, results='hide'}
suppressPackageStartupMessages({
  library(grDevices)
  library(maftools)
  library(data.table)
  library(tidyr)
  library(dplyr)
  library(stringr)
  library(GenomicRanges)
  library(readr)
  library("mclust")
  library(ggplot2)
  library(circlize)
  library(RCircos)
  library("RColorBrewer")
  library(grDevices)
  library(ComplexHeatmap)
  library(TxDb.Hsapiens.UCSC.hg38.knownGene)
  library('org.Hs.eg.db')
  library(AnnotationDbi)
  library(biovizBase)
  library(grid)
  library(purrr)
  library(ggrepel)
  library("DT")
})
suppressMessages({
  source("/hpcnfs/scratch/P_DIMA_PALL/report_ALL/scripts_report/R_project/src/utils.R")
})
```

## Setup & Preprocessing

### Genelist preprocessing

The analysis will focus only on a provided specific list of 300 genes.

All oncoplots will be filtered for this specific set of 300 genes.

```{r preprocessing_genelist, results='hide'}
samples <- list.dirs(params$maf_tumor_dir, full.names = FALSE, recursive = FALSE)

final_genetable <- bind_rows(
  read_csv(params$gene_table1),
  read_csv(params$gene_table2)
)

genelist <- unique(final_genetable$Hugo_Symbol)

id_map <- read.csv(params$id_data, stringsAsFactors = FALSE)
barcode_map <- setNames(id_map$SAMPLENAME, id_map$code)

maf_obj <- readRDS("/hpcnfs/scratch/P_DIMA_PALL/report_ALL/scripts_report/R_project/data/maf_obj.rds")
mafobj_allgenes <- readRDS("/hpcnfs/scratch/P_DIMA_PALL/report_ALL/scripts_report/R_project/data/mafobj_allgenes.rds")
cnv_mapped <- readRDS("/hpcnfs/scratch/P_DIMA_PALL/report_ALL/scripts_report/R_project/data/cnv_mapped.rds")
clinical_mapped <- readRDS("/hpcnfs/scratch/P_DIMA_PALL/report_ALL/scripts_report/R_project/data/clinical_mapped.rds")
cnv_plot_data <- readRDS("/hpcnfs/scratch/P_DIMA_PALL/report_ALL/scripts_report/R_project/data/cnv_plot_data.rds")
dragen_plot_data <- readRDS("/hpcnfs/scratch/P_DIMA_PALL/report_ALL/scripts_report/R_project/data/dragen_plot_data.rds")
cn_oncoplot_data <- readRDS("/hpcnfs/scratch/P_DIMA_PALL/report_ALL/scripts_report/R_project/data/cn_oncoplot_data.rds")
gene_cn_final_table <- readRDS("/hpcnfs/scratch/P_DIMA_PALL/report_ALL/scripts_report/R_project/data/gene_cn_final_table.rds")
```

### CNV files preprocessing

The cnv files from DRAGEN are preprocessed and joined in a single final file.

### Maf Object generation

The MAF object is generated through maftools (2.24.0) by combining the mafs from the small variants analysis (DRAGEN -> Variantalker), the cnv files from DRAGEN and the clinical data provided.

### Preprocessing MAF

The MAF object is preprocessed to:

  - add columns that contain an additional classification based on:
  
    - the most pathogenic tag of the variant between the Clinvar, Cancervar and VARSOME annotations
    
    - the most pathogenic tag of the variant between the Clinvar and Cancervar annotations
    
  - filter the variant list by:
  
    - discarding variants with VAF < 0.05
    
    - discarding fishing genes ("MUC", "TTN", "RYR", "OR", "KRT")
    
    - discarding variants in non important regions such as "Splice_Site", "Silent", "Intron", "5'UTR" ...

```{r clin_summary_setup}
clinvar_dict <- c("Pathogenic" = "Pathogenic",
    "Pathogenic/Likely_pathogenic" = "Pathogenic",
    "Likely_pathogenic" = "Likely_Pathogenic",
    "Uncertain_significance" = "Uncertain_Significance",
    "Conflicting_classifications_of_pathogenicity" = "Uncertain_Significance",
    "Conflicting_interpretations_of_pathogenicity" = "Uncertain_Significance",
    "Likely_benign" = "Likely_Benign",
    "Benign/Likely_benign" = "Likely_Benign",
    "Benign" = "Benign",
    "drug_response" = "Drug_Response",
    "not_provided" = "Not_Provided",
    "no_classifications_from_unflagged_records" = "Unknown",
    "risk_factor" = "Unknown",
    "Affects" = "Unknown",
    "no_classification_for_the_single_variant" = "Unknown",
    "association" = "Unknown",
    "other" = "Unknown",
    "Uncertain_risk_allele" = "Unknown",
    "Likely_risk_allele" = "Unknown",
    "_low_penetrance" = "Unknown",
    "protective" = "Unknown",
    "association_not_found" = "Unknown",
    "Established_risk_allele" = "Unknown",
    "confers_sensitivity" = "Unknown")

cancervar_dict <- list("Tier_I_strong" = "Pathogenic",
    "Tier_II_potential" = "Likely_Pathogenic",
    "Tier_III_Uncertain" = "Uncertain_Significance",
    "Tier_IV_benign" = "Benign")

varsome_dict  <- list("TierI" = "Pathogenic",
                      "TierIV" = "Benign",
                      "TierIII" = "Uncertain_Significance",
                      "??" = "Unknown",
                      "TierII" = "Likely_pathogenic",
                      "???" = "Unknown")

priority_order <- list("Pathogenic",
    "Likely_Pathogenic",
    "Uncertain_Significance",
    "Likely_Benign",
    "Benign",
    "Drug_Response",
    "Not_Provided",
    "Unknown")

cols_to_check_clin_canc <- list("CancerVar", "ClinVar_VCF_CLNSIG")
cols_to_check_chiara <- list("VARSOME", "CancerVar", "ClinVar_VCF_CLNSIG")
```

### Genelists filtering

For convenience, a series of maf objects are generated by subsetting the original maf to keep only the variants with a pathogenicity tier above a certain tier. This will be useful to generate the final oncoplots.

The list of genes associated with each maf is also generated.

```{r genelists, results='hide'}
maf_300 <- mafobj_allgenes@data[Hugo_Symbol %in% genelist, ]
cnv_300 <- cnv_mapped[cnv_mapped$Hugo_Symbol %in% genelist, ]
mafobj_300 <- read.maf(maf = maf_300, clinicalData = clinical_mapped, cnTable = cnv_300)

tier1 <- c("Pathogenic")
tier2 <- c("Pathogenic", "Likely_Pathogenic")
tier3 <- c("Pathogenic", "Likely_Pathogenic", "Uncertain_Significance")

maf_t1 <- maf_300[maf_300$Clinical_Significance_Summary_Clinvar_Cancervar %in% tier1, ]
cnv_t1 <- cnv_300[cnv_300$Classification %in% tier1, ]
mafobj_t1 <- read.maf(maf = maf_t1, cnTable = cnv_t1, clinicalData = clinical_mapped)

maf_t2 <- maf_300[maf_300$Clinical_Significance_Summary_Clinvar_Cancervar %in% tier2, ]
cnv_t2 <- cnv_300[cnv_300$Classification %in% tier2, ]
mafobj_t2 <- read.maf(maf = maf_t2, cnTable = cnv_t2, clinicalData = clinical_mapped)

maf_t3 <- maf_300[maf_300$Clinical_Significance_Summary_Clinvar_Cancervar %in% tier3, ]
cnv_t3 <- cnv_300[cnv_300$Classification %in% tier3, ]
mafobj_t3 <- read.maf(maf = maf_t3, cnTable = cnv_t3, clinicalData = clinical_mapped)

maf_t1_c <- maf_300[maf_300$Clinical_Significance_Summary_Chiara %in% tier1, ]
cnv_t1_c <- cnv_300[cnv_300$Classification %in% tier1, ]
mafobj_t1_c <- read.maf(maf = maf_t1_c, cnTable = cnv_t1_c, clinicalData = clinical_mapped)

maf_t2_c <- maf_300[maf_300$Clinical_Significance_Summary_Chiara %in% tier2, ]
cnv_t2_c <- cnv_300[cnv_300$Classification %in% tier2, ]
mafobj_t2_c <- read.maf(maf = maf_t2_c, cnTable = cnv_t2_c, clinicalData = clinical_mapped)

maf_t3_c <- maf_300[maf_300$Clinical_Significance_Summary_Chiara %in% tier3, ]
cnv_t3_c <- cnv_300[cnv_300$Classification %in% tier3, ]
mafobj_t3_c <- read.maf(maf = maf_t3_c, cnTable = cnv_t3_c, clinicalData = clinical_mapped)

genes_t1   <- getGeneSummary(mafobj_t1)$Hugo_Symbol
genes_t2   <- getGeneSummary(mafobj_t2)$Hugo_Symbol
genes_t3   <- getGeneSummary(mafobj_t3)$Hugo_Symbol

genes_t1_c <- getGeneSummary(mafobj_t1_c)$Hugo_Symbol
genes_t2_c <- getGeneSummary(mafobj_t2_c)$Hugo_Symbol
genes_t3_c <- getGeneSummary(mafobj_t3_c)$Hugo_Symbol
```

```{r filters_genes, results='hide'}
clin_data <- mafobj_allgenes@clinical.data
clin_data_sorted <- clin_data[order(clin_data$T1221, clin_data$Sample_group), ]
clin_data_sorted_samples_paired <- clin_data[order(clin_data$Tumor_Sample_Barcode, clin_data$T1221, clin_data$Sample_group), ]
clin_data_sorted[, patient_id := sub("-.*", "", Tumor_Sample_Barcode)]
clin_data_sorted[, sample_suffix := sub(".*-(.*)", "\\1", Tumor_Sample_Barcode)]
clin_data_sorted[, sample_num := as.integer(sub("ALL([0-9]+).*", "\\1", Tumor_Sample_Barcode))]

sample_order <- clin_data_sorted$Tumor_Sample_Barcode
sample_order_paired <- clin_data_sorted_samples_paired$Tumor_Sample_Barcode

# gene filters
topN <- params$num_top_genes

top_t1   <- head(mafobj_t1@gene.summary[order(-MutatedSamples)]$Hugo_Symbol, topN)
top_t2   <- head(mafobj_t2@gene.summary[order(-MutatedSamples)]$Hugo_Symbol, topN)
top_t3   <- head(mafobj_t3@gene.summary[order(-MutatedSamples)]$Hugo_Symbol, topN)

top_t1_c <- head(mafobj_t1_c@gene.summary[order(-MutatedSamples)]$Hugo_Symbol, topN)
top_t2_c <- head(mafobj_t2_c@gene.summary[order(-MutatedSamples)]$Hugo_Symbol, topN)
top_t3_c <- head(mafobj_t3_c@gene.summary[order(-MutatedSamples)]$Hugo_Symbol, topN)
```

### Sample lists filtering

Similarly to the genes, the sample list is also filtered by clinical conditions and ordered to be used in the final oncoplots.

```{r filters_samples, results='hide'}
# sample filters
sample_order_diag_relapse <- clin_data_sorted[Sample_group %in% c("diag", "relapse1")][order(T1221, sample_num, sample_suffix)]$Tumor_Sample_Barcode
maf_obj_filtered_diag_relapse <- subsetMaf(maf = mafobj_allgenes, tsb = sample_order_diag_relapse)

samples_neg_diag_relapse <- clin_data_sorted[T1221 == "NEG" & Sample_group %in% c("diag", "relapse1")][order(sample_num, sample_suffix)]$Tumor_Sample_Barcode

samples_pos_diag_relapse <- clin_data_sorted[T1221 == "POS" & Sample_group %in% c("diag", "relapse1")][order(sample_num, sample_suffix)]$Tumor_Sample_Barcode


sample_order_diag <- clin_data_sorted[Sample_group == "diag"][order(sample_num, sample_suffix)]$Tumor_Sample_Barcode
maf_obj_filtered_diag <- subsetMaf(maf = mafobj_allgenes, tsb = sample_order_diag)

sample_order_relapse <- clin_data_sorted[Sample_group == "relapse1"][order(sample_num, sample_suffix)]$Tumor_Sample_Barcode
maf_obj_filtered_relapse <- subsetMaf(maf = mafobj_allgenes, tsb = sample_order_relapse)

samples_neg <- clin_data_sorted[T1221 == "NEG"][order(sample_num, sample_suffix)]$Tumor_Sample_Barcode
maf_obj_filtered_neg <- subsetMaf(maf = mafobj_allgenes, tsb = samples_neg)
samples_pos <- clin_data_sorted[T1221 == "POS"][order(sample_num, sample_suffix)]$Tumor_Sample_Barcode
maf_obj_filtered_pos <- subsetMaf(maf = mafobj_allgenes, tsb = samples_pos)

sample_order_patient_adj <- clin_data_sorted[order(T1221, sample_num, sample_suffix)]$Tumor_Sample_Barcode
```

```{r setup_directories}
# setup directories
output_dir <- file.path(params$results_dir, paste0(params$manual_timestamp, "_", params$project))
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}
```

## Summary

The summary plots were generated with maftools.

They provide an overview of the cohort data by showing informations about the variant classification, the top 10 mutated genes and the distribution across samples. 

In the top 10 mutated genes section:

  - the bar length represents the total number of mutations found for that particular gene across all samples
  
  - the percentage represents the fraction of samples with at least one mutation in that gene

### Summary - all samples

```{r summary1}
plotmafSummary(maf = mafobj_allgenes, rmOutlier = TRUE, addStat = "median", dashboard = TRUE, titvRaw = FALSE)
```

### Summary - diagnosis samples

```{r summary2}
plotmafSummary(maf = maf_obj_filtered_diag, rmOutlier = TRUE, addStat = 'median', dashboard = TRUE, titvRaw = FALSE)
```

### Summary - first relapse samples

```{r summary3}
plotmafSummary(maf = maf_obj_filtered_relapse, rmOutlier = TRUE, addStat = 'median', dashboard = TRUE, titvRaw = FALSE)
```

## Oncoplots

All small variants oncoplots were generated with maftools.

Each oncoplot was generated by taking into account:

  - The tier filter (above I only, above II only and above III only) and the tiering method (most pathogenic across Clinvar, Cancervar and VARSOME (VCC) vs Clinvar and Cancervar only (CC))
  
  - The number of top genes to plot (25)
  
  - The sample grouping (either by clinical info (T1221 = based on detection of the traslocation of chr 12 to chr 21, Sample_group = diagnosis, relapses), or by patient (diagnosis and relapses grouped together))

### Most pathogenic in VCC, Tier I and II genes

#### All samples

```{r oncoplot1, results='hide', echo=FALSE}
main_onco <- oncoplot(
  maf = mafobj_t2_c,
  genes = top_t2_c,
  showTumorSampleBarcodes = TRUE, 
  barcode_mar = 8, 
  clinicalFeatures = c("T1221", "Sample_group"), 
  SampleNamefontSize = 1.3,
  sampleOrder = sample_order
  )
gene_order <- rownames(main_onco$oncomatrix)
print(main_onco)
```


#### All samples, by patient

```{r oncoplot1bypat, results='hide', echo=FALSE}
oncoplot(
  maf = mafobj_t2_c,
  genes = gene_order, 
  keepGeneOrder = TRUE,
  GeneOrderSort = FALSE,
  showTumorSampleBarcodes = TRUE, 
  barcode_mar = 12.5, 
  clinicalFeatures = c("T1221", "Sample_group"), 
  SampleNamefontSize = 1.3,
  sampleOrder = sample_order_patient_adj
  )
```

#### Positive samples

```{r oncoplot1pos, results='hide', echo=FALSE}
oncoplot(
  maf = subsetMaf(maf = mafobj_t2_c, tsb = samples_pos),
  genes = gene_order, 
  keepGeneOrder = TRUE,
  GeneOrderSort = FALSE,
  showTumorSampleBarcodes = TRUE, 
  barcode_mar = 12.5, 
  clinicalFeatures = c("T1221", "Sample_group"), 
  SampleNamefontSize = 1.3,
  sampleOrder = samples_pos
  )
```

#### Negative samples

```{r oncoplot1neg, results='hide', echo=FALSE}
oncoplot(
  maf = subsetMaf(maf = mafobj_t2_c, tsb = samples_neg),
  genes = gene_order, 
  keepGeneOrder = TRUE,
  GeneOrderSort = FALSE,
  showTumorSampleBarcodes = TRUE, 
  barcode_mar = 12.5, 
  clinicalFeatures = c("T1221", "Sample_group"), 
  SampleNamefontSize = 1.3,
  sampleOrder = samples_neg
  )
```




### Most pathogenic in VCC, Tier I, II and III genes

#### All samples

```{r oncoplot2, results='hide', echo=FALSE}
main_onco <- oncoplot(
  maf = mafobj_t3_c,
  genes = top_t3_c, 
  showTumorSampleBarcodes = TRUE, 
  barcode_mar = 12.5, 
  clinicalFeatures = c("T1221", "Sample_group"), 
  SampleNamefontSize = 1.3,
  sampleOrder = sample_order
  )
gene_order <- rownames(main_onco$oncomatrix)
print(main_onco)
```

#### All samples, by patient

```{r oncoplot2bypat, results='hide', echo=FALSE}
oncoplot(
  maf = mafobj_t3_c,
  genes = gene_order, 
  keepGeneOrder = TRUE,
  GeneOrderSort = FALSE,
  showTumorSampleBarcodes = TRUE, 
  barcode_mar = 12.5, 
  clinicalFeatures = c("T1221", "Sample_group"), 
  SampleNamefontSize = 1.3,
  sampleOrder = sample_order_patient_adj
  )
```

#### Positive samples

```{r oncoplot2pos, results='hide', echo=FALSE}
oncoplot(
  maf = subsetMaf(maf = mafobj_t3_c, tsb = samples_pos),
  genes = gene_order, 
  keepGeneOrder = TRUE,
  GeneOrderSort = FALSE,
  showTumorSampleBarcodes = TRUE, 
  barcode_mar = 12.5, 
  clinicalFeatures = c("T1221", "Sample_group"), 
  SampleNamefontSize = 1.3,
  sampleOrder = samples_pos
  )
```

#### Negative samples

```{r oncoplot2neg, results='hide', echo=FALSE}
oncoplot(
  maf = subsetMaf(maf = mafobj_t3_c, tsb = samples_neg),
  genes = gene_order, 
  keepGeneOrder = TRUE,
  GeneOrderSort = FALSE,
  showTumorSampleBarcodes = TRUE, 
  barcode_mar = 12.5, 
  clinicalFeatures = c("T1221", "Sample_group"), 
  SampleNamefontSize = 1.3,
  sampleOrder = samples_neg
  )
```



### Most pathogenic in CC, Tier I and II genes

#### All samples

```{r oncoplot3, results='hide', echo=FALSE}
main_onco <- oncoplot(
  maf = mafobj_t2,
  genes = top_t2, 
  showTumorSampleBarcodes = TRUE, 
  barcode_mar = 12.5, 
  clinicalFeatures = c("T1221", "Sample_group"), 
  SampleNamefontSize = 1.3,
  sampleOrder = sample_order
  )
gene_order <- rownames(main_onco$oncomatrix)
print(main_onco)
```

#### All samples, by patient

```{r oncoplot3bypat, results='hide', echo=FALSE}
oncoplot(
  maf = mafobj_t2,
  genes = gene_order, 
  keepGeneOrder = TRUE,
  GeneOrderSort = FALSE,
  showTumorSampleBarcodes = TRUE, 
  barcode_mar = 12.5, 
  clinicalFeatures = c("T1221", "Sample_group"), 
  SampleNamefontSize = 1.3,
  sampleOrder = sample_order_patient_adj
  )
```

#### Positive samples

```{r oncoplot3pos, results='hide', echo=FALSE}
oncoplot(
  maf = subsetMaf(maf = mafobj_t2, tsb = samples_pos),
  genes = gene_order, 
  keepGeneOrder = TRUE,
  GeneOrderSort = FALSE,
  showTumorSampleBarcodes = TRUE, 
  barcode_mar = 12.5, 
  clinicalFeatures = c("T1221", "Sample_group"), 
  SampleNamefontSize = 1.3,
  sampleOrder = samples_pos
  )
```

#### Negative samples

```{r oncoplot3neg, results='hide', echo=FALSE}
oncoplot(
  maf = subsetMaf(maf = mafobj_t2, tsb = samples_neg),
  genes = gene_order, 
  keepGeneOrder = TRUE,
  GeneOrderSort = FALSE,
  showTumorSampleBarcodes = TRUE, 
  barcode_mar = 12.5, 
  clinicalFeatures = c("T1221", "Sample_group"), 
  SampleNamefontSize = 1.3,
  sampleOrder = samples_neg
  )
```





### Most pathogenic in CC, Tier I, II and III genes, all samples

#### All samples

```{r oncoplot4, results='hide', echo=FALSE}
main_onco <- oncoplot(
  maf = mafobj_t3,
  genes = top_t3, 
  showTumorSampleBarcodes = TRUE, 
  barcode_mar = 12.5, 
  clinicalFeatures = c("T1221", "Sample_group"), 
  SampleNamefontSize = 1.3,
  sampleOrder = sample_order
  )
gene_order <- rownames(main_onco$oncomatrix)
print(main_onco)
```

#### All samples, by patient

```{r oncoplot4bypat, results='hide', echo=FALSE}
oncoplot(
  maf = mafobj_t3,
  genes = gene_order, 
  keepGeneOrder = TRUE,
  GeneOrderSort = FALSE,
  showTumorSampleBarcodes = TRUE, 
  barcode_mar = 12.5, 
  clinicalFeatures = c("T1221", "Sample_group"), 
  SampleNamefontSize = 1.3,
  sampleOrder = sample_order_patient_adj
  )
```

#### Positive samples

```{r oncoplot4pos, results='hide', echo=FALSE}
oncoplot(
  maf = subsetMaf(maf = mafobj_t3, tsb = samples_pos),
  genes = gene_order, 
  keepGeneOrder = TRUE,
  GeneOrderSort = FALSE,
  showTumorSampleBarcodes = TRUE, 
  barcode_mar = 12.5, 
  clinicalFeatures = c("T1221", "Sample_group"), 
  SampleNamefontSize = 1.3,
  sampleOrder = samples_pos
  )
```

#### Negative samples

```{r oncoplot4neg, results='hide', echo=FALSE}
oncoplot(
  maf = subsetMaf(maf = mafobj_t3, tsb = samples_neg),
  genes = gene_order, 
  keepGeneOrder = TRUE,
  GeneOrderSort = FALSE,
  showTumorSampleBarcodes = TRUE, 
  barcode_mar = 12.5, 
  clinicalFeatures = c("T1221", "Sample_group"), 
  SampleNamefontSize = 1.3,
  sampleOrder = samples_neg
  )
```




### Most pathogenic in VCC, Tier I and II genes, only diagnosis and relapse 1 samples

#### All samples

```{r oncoplot5, results='hide', echo=FALSE}
main_onco <- oncoplot(
  maf = subsetMaf(maf = mafobj_t2_c, tsb = sample_order_diag_relapse),
  genes = top_t2_c, 
  showTumorSampleBarcodes = TRUE, 
  barcode_mar = 12.5, 
  clinicalFeatures = c("T1221", "Sample_group"), 
  SampleNamefontSize = 1.3,
  sampleOrder = sample_order
  )
gene_order <- rownames(main_onco$oncomatrix)
print(main_onco)
```

#### All samples, by patient

```{r oncoplot5bypat, results='hide', echo=FALSE}
oncoplot(
  maf = subsetMaf(maf = mafobj_t2_c, tsb = sample_order_diag_relapse),
  genes = gene_order, 
  keepGeneOrder = TRUE,
  GeneOrderSort = FALSE,
  showTumorSampleBarcodes = TRUE, 
  barcode_mar = 12.5, 
  clinicalFeatures = c("T1221", "Sample_group"), 
  SampleNamefontSize = 1.3,
  sampleOrder = sample_order_patient_adj
  )
```

#### Positive samples

```{r oncoplot5pos, results='hide', echo=FALSE}
oncoplot(
  maf = subsetMaf(maf = mafobj_t2_c, tsb = samples_pos_diag_relapse),
  genes = gene_order, 
  keepGeneOrder = TRUE,
  GeneOrderSort = FALSE,
  showTumorSampleBarcodes = TRUE, 
  barcode_mar = 12.5, 
  clinicalFeatures = c("T1221", "Sample_group"), 
  SampleNamefontSize = 1.3,
  sampleOrder = samples_pos
  )
```

#### Negative samples

```{r oncoplot5neg, results='hide', echo=FALSE}
oncoplot(
  maf = subsetMaf(maf = mafobj_t2_c, tsb = samples_neg_diag_relapse),
  genes = gene_order, 
  keepGeneOrder = TRUE,
  GeneOrderSort = FALSE,
  showTumorSampleBarcodes = TRUE, 
  barcode_mar = 12.5, 
  clinicalFeatures = c("T1221", "Sample_group"), 
  SampleNamefontSize = 1.3,
  sampleOrder = samples_neg
  )
```




### Most pathogenic in VCC, Tier I, II and III genes, only diagnosis and relapse 1 samples

#### All samples

```{r oncoplot6, results='hide', echo=FALSE}
main_onco <- oncoplot(
  maf = subsetMaf(maf = mafobj_t3_c, tsb = sample_order_diag_relapse),
  genes = top_t3_c, 
  showTumorSampleBarcodes = TRUE, 
  barcode_mar = 12.5, 
  clinicalFeatures = c("T1221", "Sample_group"), 
  SampleNamefontSize = 1.3,
  sampleOrder = sample_order
  )
gene_order <- rownames(main_onco$oncomatrix)
print(main_onco)
```

#### All samples, by patient

```{r oncoplot6bypat, results='hide', echo=FALSE}
oncoplot(
  maf = subsetMaf(maf = mafobj_t3_c, tsb = sample_order_diag_relapse),
  genes = gene_order, 
  keepGeneOrder = TRUE,
  GeneOrderSort = FALSE,
  showTumorSampleBarcodes = TRUE, 
  barcode_mar = 12.5, 
  clinicalFeatures = c("T1221", "Sample_group"), 
  SampleNamefontSize = 1.3,
  sampleOrder = sample_order_patient_adj
  )
```

#### Positive samples

```{r oncoplot6pos, results='hide', echo=FALSE}
oncoplot(
  maf = subsetMaf(maf = mafobj_t3_c, tsb = samples_pos_diag_relapse),
  genes = gene_order, 
  keepGeneOrder = TRUE,
  GeneOrderSort = FALSE,
  showTumorSampleBarcodes = TRUE, 
  barcode_mar = 12.5, 
  clinicalFeatures = c("T1221", "Sample_group"), 
  SampleNamefontSize = 1.3,
  sampleOrder = samples_pos
  )
```

#### Negative samples

```{r oncoplot6neg, results='hide', echo=FALSE}
oncoplot(
  maf = subsetMaf(maf = mafobj_t3_c, tsb = samples_neg_diag_relapse),
  genes = gene_order, 
  keepGeneOrder = TRUE,
  GeneOrderSort = FALSE,
  showTumorSampleBarcodes = TRUE, 
  barcode_mar = 12.5, 
  clinicalFeatures = c("T1221", "Sample_group"), 
  SampleNamefontSize = 1.3,
  sampleOrder = samples_neg
  )
```




### Most pathogenic in CC, Tier I, and II genes, only diagnosis and relapse 1 samples

#### All samples

```{r oncoplot7, results='hide', echo=FALSE}
main_onco <- oncoplot(
  maf = subsetMaf(maf = mafobj_t2, tsb = sample_order_diag_relapse),
  genes = top_t2, 
  showTumorSampleBarcodes = TRUE, 
  barcode_mar = 12.5, 
  clinicalFeatures = c("T1221", "Sample_group"), 
  SampleNamefontSize = 1.3,
  sampleOrder = sample_order
  )
gene_order <- rownames(main_onco$oncomatrix)
print(main_onco)
```

#### All samples, by patient

```{r oncoplot7bypat, results='hide', echo=FALSE}
oncoplot(
  maf = subsetMaf(maf = mafobj_t2, tsb = sample_order_diag_relapse),
  genes = gene_order, 
  keepGeneOrder = TRUE,
  GeneOrderSort = FALSE,
  showTumorSampleBarcodes = TRUE, 
  barcode_mar = 12.5, 
  clinicalFeatures = c("T1221", "Sample_group"), 
  SampleNamefontSize = 1.3,
  sampleOrder = sample_order_patient_adj
  )
```

#### Positive samples

```{r oncoplot7pos, results='hide', echo=FALSE}
oncoplot(
  maf = subsetMaf(maf = mafobj_t2, tsb = samples_pos_diag_relapse),
  genes = gene_order, 
  keepGeneOrder = TRUE,
  GeneOrderSort = FALSE,
  showTumorSampleBarcodes = TRUE, 
  barcode_mar = 12.5, 
  clinicalFeatures = c("T1221", "Sample_group"), 
  SampleNamefontSize = 1.3,
  sampleOrder = samples_pos
  )
```

#### Negative samples

```{r oncoplot7neg, results='hide', echo=FALSE}
oncoplot(
  maf = subsetMaf(maf = mafobj_t2, tsb = samples_neg_diag_relapse),
  genes = gene_order, 
  keepGeneOrder = TRUE,
  GeneOrderSort = FALSE,
  showTumorSampleBarcodes = TRUE, 
  barcode_mar = 12.5, 
  clinicalFeatures = c("T1221", "Sample_group"), 
  SampleNamefontSize = 1.3,
  sampleOrder = samples_neg
  )
```




### Most pathogenic in CC, Tier I, II and III genes, only diagnosis and relapse 1 samples

#### All samples

```{r oncoplot8, results='hide', echo=FALSE}
main_onco <- oncoplot(
  maf = subsetMaf(maf = mafobj_t3, tsb = sample_order_diag_relapse),
  genes = top_t3, 
  showTumorSampleBarcodes = TRUE, 
  barcode_mar = 12.5, 
  clinicalFeatures = c("T1221", "Sample_group"), 
  SampleNamefontSize = 1.3,
  sampleOrder = sample_order
  )
gene_order <- rownames(main_onco$oncomatrix)
print(main_onco)
```

#### All samples, by patient

```{r oncoplot8bypat, results='hide', echo=FALSE}
oncoplot(
  maf = subsetMaf(maf = mafobj_t3, tsb = sample_order_diag_relapse),
  genes = gene_order, 
  keepGeneOrder = TRUE,
  GeneOrderSort = FALSE,
  showTumorSampleBarcodes = TRUE, 
  barcode_mar = 12.5, 
  clinicalFeatures = c("T1221", "Sample_group"), 
  SampleNamefontSize = 1.3,
  sampleOrder = sample_order_patient_adj
  )
```

#### Positive samples

```{r oncoplot8pos, results='hide', echo=FALSE}
oncoplot(
  maf = subsetMaf(maf = mafobj_t3, tsb = samples_pos_diag_relapse),
  genes = gene_order, 
  keepGeneOrder = TRUE,
  GeneOrderSort = FALSE,
  showTumorSampleBarcodes = TRUE, 
  barcode_mar = 12.5, 
  clinicalFeatures = c("T1221", "Sample_group"), 
  SampleNamefontSize = 1.3,
  sampleOrder = samples_pos
  )
```

#### Negative samples

```{r oncoplot8neg, results='hide', echo=FALSE}
oncoplot(
  maf = subsetMaf(maf = mafobj_t3, tsb = samples_neg_diag_relapse),
  genes = gene_order, 
  keepGeneOrder = TRUE,
  GeneOrderSort = FALSE,
  showTumorSampleBarcodes = TRUE, 
  barcode_mar = 12.5, 
  clinicalFeatures = c("T1221", "Sample_group"), 
  SampleNamefontSize = 1.3,
  sampleOrder = samples_neg
  )
```



## Other Plots

Additional secondary plots from maftools.

### Transitions-Transversions

The plot contains informations on the sample rate of transition vs transversions for all samples and their distribution across samples.

```{r transition_transversions, fig.height=13}
maf_obj.titv = titv(maf = mafobj_allgenes, plot = FALSE, useSyn = TRUE)
plotTiTv(res = maf_obj.titv, showBarcodes = TRUE)
```

### TMB Comparison Graph

The plot shows a comparison between the analyzed cohort and a set of standardized cohorts from the literature and tumor classification. The cohorts are compared by average Tumor mutational burden.

```{r TMB_comparison}
maf_obj.mutload = tcgaCompare(maf = mafobj_allgenes, cohortName = 'PALL', logscale = TRUE, capture_size = 50)
```

### VAF Plots

The VAF plots show the VAF distribution of the top 20 genes in the selected 300 gene list by VAF.

#### VAF Plot, positive samples, most pathogenic in CC, Tier I and II genes

```{r vaf_plot1, results='hide', echo=FALSE}
plotVaf(maf = subsetMaf(maf = mafobj_t2, tsb = samples_pos), vafCol = 'tumor_f', genes = genes_t2)
```

#### VAF Plot, negative samples, most pathogenic in CC, Tier I and II genes

```{r vaf_plot2, results='hide', echo=FALSE}
plotVaf(maf = subsetMaf(maf = mafobj_t2, tsb = samples_neg), vafCol = 'tumor_f', genes = genes_t2)
```

#### VAF Plot, positive samples, most pathogenic in VCC, Tier I and II genes

```{r vaf_plot3, results='hide', echo=FALSE}
plotVaf(maf = subsetMaf(maf = mafobj_t2_c, tsb = samples_pos), vafCol = 'tumor_f', genes = genes_t2_c)
```

#### VAF Plot, negative samples, most pathogenic in VCC, Tier I and II genes

```{r vaf_plot4, results='hide', echo=FALSE}
plotVaf(maf = subsetMaf(maf = mafobj_t2_c, tsb = samples_neg), vafCol = 'tumor_f', genes = genes_t2_c)
```

## Copy number Manhattan Plots

This section contains the Manhattan plots generated from the copy number data from DRAGEN and CNVkit. 

The plots are generated with ggplot2 (4.0.0).

```{r setup_genelist_cnvkit}
gene_entrez <- AnnotationDbi::select(org.Hs.eg.db,
                                    keys = genelist,
                                    keytype = "SYMBOL",
                                    columns = c("ENTREZID"))

gene_entrez <- gene_entrez %>% filter(!is.na(ENTREZID))
txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene
gene_ranges <- genes(txdb, columns = c("gene_id"))
gene_ranges_filtered <- gene_ranges[names(gene_ranges) %in% gene_entrez$ENTREZID]
gene_df <- as.data.frame(gene_ranges_filtered) %>%
  dplyr::select(seqnames, start, end, width, strand, gene_id)

gene_annotated <- gene_entrez %>%
  inner_join(gene_df, by = c("ENTREZID" = "gene_id")) %>%
  dplyr::rename(
    Chromosome = seqnames,
    Start_Position = start,
    End_Position = end
  )
gene_annotated <- gene_annotated %>%
  mutate(Chromosome = gsub("^chr", "", Chromosome))
```

### CN CNVKit, Manhattan Plot, Segments

The CNVkit Manhattan plots have been generated from the CNVkit output which was in turn generated from the tumor and normal cram files from DRAGEN.

These plots show on the Y axis the log2R of the copy number of each region calculated directly by CNVkit. 

The plot has been centered at value log2R = 0 for all patients and the log2R values have been capped at +- 2.5 to limit the outliers effect on influencing the plot scaling from patient to patient. 

An additional gene label has been added by taking the initial list of 300 genes and evaluating if each one of these genes would fall in a copy number region with a log2R absolute value > 0.5.

```{r manhattan_manual_seg_cnvkit}
plot_list <- list()
plot_params <- cnv_plot_data$metadata

for (sample in names(cnv_plot_data$samples)) {

  cns3 <- cnv_plot_data$samples[[sample]]$cns3
  centers <- cnv_plot_data$samples[[sample]]$centers
  df_annot_filtered <- cnv_plot_data$samples[[sample]]$df_annot_filtered

  n_chrom <- length(unique(cns3$chromosome))
  chrom_colors <- colorRampPalette(
    brewer.pal(12, "Paired")
  )(n_chrom)

  x_total <- max(cns3$pos_cum) - min(cns3$pos_cum)
  nudge_range <- 0.05 * x_total

  p <- ggplot(
    cns3,
    aes(
      x = pos_cum,
      y = log2_capped,
      group = chromosome
    )
  ) +
    geom_segment(
      aes(
        x = start + chr_cum_start,
        xend = end + chr_cum_start,
        y = log2_capped,
        yend = log2_capped,
        color = chromosome
      ),
      linewidth = 1
    ) +
    geom_hline(
      yintercept = 0,
      linetype = "dashed",
      color = "red"
    ) +
    geom_text_repel(
      data = df_annot_filtered,
      aes(
        x = pos_cum,
        y = log2_capped,
        label = genes
      ),
      size = 2,
      nudge_y = ifelse(
        df_annot_filtered$log2_capped > 0,
        0.2,
        -0.2
      ),
      nudge_x = runif(
        nrow(df_annot_filtered),
        -nudge_range,
        nudge_range
      ),
      direction = "y",
      segment.size = 0.2,
      segment.color = "grey60",
      box.padding = 0.2,
      force = 2,
      force_pull = 0.5
    ) +
    scale_y_continuous(
      limits = c(
        plot_params$log2_cap_min,
        plot_params$log2_cap_max
      )
    ) +
    scale_x_continuous(
      breaks = centers$center,
      labels = centers$chromosome,
      expand = c(0,0)
    ) +
    theme_bw() +
    theme(
      axis.text.x = element_text(angle = 90, size = 6),
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank(),
      aspect.ratio = plot_params$aspect_ratio
    ) +
    labs(
      title = paste(
        "Copy number segments (line plot) for sample",
        sample
      ),
      x = "Chromosome",
      y = "Log2 copy number ratio"
    )

  plot_list[[sample]] <- p

}
```

```{r results='asis', echo=FALSE}
patient_ids <- unique(sub("-.*", "", names(plot_list)))

for (patient in patient_ids) {
  cat(sprintf(
    "<details style='margin-bottom:12px;'>\n<summary><b>%s</b></summary>\n\n",
    patient
  ))
  samples <- grep(
    paste0("^", patient, "-"),
    names(plot_list),
    value = TRUE
  )
  for (sample in samples) {
    cat(sprintf("<h4>%s</h4>\n", sample))
    print(plot_list[[sample]])
  }
  cat("</details>\n\n")
}
```

### CN DRAGEN annotation files + CNVKit seg files, Manhattan Plot, Segments

The DRAGEN Manhattan plots have been generated directly from the DRAGEN output that has been annotated by Variantalker (ClassifyCNV, ".cnv.annotated.tsv" files).

These plots show on the Y axis the log2R of the copy number of each region calculated directly by CNVkit.

The plot has been centered at value log2R = 0 for all patients and the log2R values have been capped at +- 2.5 to limit the outliers effect on influencing the plot scaling from patient to patient. 

An additional gene label has been added by taking the list of genes contained in the column "Known or predicted dosage-sensitive genes" from each .cnv.annotated.tsv patient file.

Only the mutations classified as Pathogenic or Likely Pathogenic by ClassifyCNV were taken.

In this case the threshold has been removed since the genes listed in the annotated file should be already above threshold, considering they have been annotated for copy number effects.

```{r plot_dragen_manhattan}
plot_params <- dragen_plot_data$metadata
plot_list <- list()

for (sample in names(dragen_plot_data$samples)) {

  cns2 <- dragen_plot_data$samples[[sample]]$cns2
  centers <- dragen_plot_data$samples[[sample]]$centers
  df_annot <- dragen_plot_data$samples[[sample]]$df_annot

  n_chrom <- length(unique(cns2$chromosome))

  chrom_colors <- setNames(
    colorRampPalette(
      RColorBrewer::brewer.pal(12, "Paired")
    )(n_chrom),
    sort(unique(cns2$chromosome))
  )

  x_total <- max(cns2$pos_cum) - min(cns2$pos_cum)
  nudge_range <- 0.08 * x_total

  p <- ggplot(
    cns2,
    aes(
      x = pos_cum,
      y = log2_capped,
      group = chromosome
    )
  ) +
    geom_segment(
      aes(
        x = start + chr_cum_start,
        xend = end + chr_cum_start,
        y = log2_capped,
        yend = log2_capped,
        color = chromosome
      ),
      linewidth = 1
    ) +
    geom_hline(
      yintercept = 0,
      linetype = "dashed",
      color = "red"
    ) +
    scale_color_manual(values = chrom_colors) +
    scale_x_continuous(
      breaks = centers$center,
      labels = centers$chromosome,
      expand = c(0,0)
    ) +
    coord_cartesian(
      ylim = c(
        plot_params$y_min,
        plot_params$y_max
      )
    ) +
    theme_bw() +
    theme(
      axis.text.x = element_text(
        angle = 90,
        size = 6
      ),
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank(),
      aspect.ratio = plot_params$aspect_ratio
    ) +
    labs(
      title = paste(
        "Copy number segments (line plot) for sample",
        sample
      ),
      x = "Chromosome",
      y = "Log2 copy number ratio"
    )

  if (!is.null(df_annot) && nrow(df_annot) > 0) {

    p <- p +
      geom_text_repel(
        data = df_annot,
        aes(
          x = pos_cum,
          y = log2_capped,
          label = genes
        ),
        size = 2,
        nudge_y = ifelse(
          df_annot$log2_capped > 0,
          0.2,
          -0.2
        ),
        nudge_x = runif(
          nrow(df_annot),
          -nudge_range,
          nudge_range
        ),
        direction = "y",
        segment.size = 0.2,
        segment.color = "grey60",
        box.padding = 0.2,
        force = 2,
        force_pull = 0.5
      )

  }
  plot_list[[sample]] <- p
}
```

```{r results='asis', echo=FALSE}
patient_ids <- unique(sub("-.*", "", names(plot_list)))

for (patient in patient_ids) {
  cat(sprintf(
    "<details style='margin-bottom:15px;'>\n<summary><b>%s</b></summary>\n\n",
    patient
  ))
  samples <- grep(
    paste0("^", patient, "-"),
    names(plot_list),
    value = TRUE
  )
  for (sample in samples) {
    cat(sprintf("<h4>%s</h4>\n", sample))
    print(plot_list[[sample]])
  }
  cat("</details>\n\n")
}
```

### CN Oncoplot, DRAGEN annotation/seg files

The copy number oncoplots have been generated from the same DRAGEN annotation files used for the DRAGEN Manhattan plots. 

3 different oncoplots have been generated based on the mutations tiering described above (tier I and above, tier II and above, tier III and above).

The logRatio used to define amplifications or deletions was already calculated by DRAGEN and the categories used to classify each copy number are simply Amplifications and Deletions.

``` {r oncoplot_setup}
all_samples <- cn_oncoplot_data$all_samples
clinical_sample_info <- cn_oncoplot_data$clinical_sample_info

t1221_colors <- c(
  NEG = "#1f78b4",
  POS = "#33a02c",
  NE  = "#b2df8a"
)

sample_group_colors <- c(
  diag = "#e41a1c",
  relapse1 = "#377eb8",
  relapse2 = "#4daf4a"
)

get_type <- function(x) strsplit(x, ";")[[1]]
```

#### CN Oncoplot, all samples, tier I

```{r oncoplot_dragen_cnbands_filter1, fig.width = 12}
plot_cn_oncoprint(
  cn_block = cn_oncoplot_data$tier_I,
  title = paste(
    "Top",
    cn_oncoplot_data$metadata$top_regions,
    "CNV Regions (cytobands × sample)"
  ),
  remove_empty_columns = FALSE
)
```

#### CN Oncoplot, all samples, tier I and II

```{r oncoplot_dragen_cnbands_filter2, fig.width = 12}
plot_cn_oncoprint(
  cn_block = cn_oncoplot_data$tier_I_II,
  title = paste(
    "Top",
    cn_oncoplot_data$metadata$top_regions,
    "CNV Regions (cytobands × sample)"
  ),
  remove_empty_columns = TRUE
)
```

#### CN Oncoplot, all samples, tier I, II and III

```{r oncoplot_dragen_cnbands_filter3, fig.width = 12}
plot_cn_oncoprint(
  cn_block = cn_oncoplot_data$tier_I_II_III,
  title = paste(
    "Top",
    cn_oncoplot_data$metadata$top_regions,
    "CNV Regions (cytobands × sample)"
  ),
  remove_empty_columns = TRUE
)
```

## CN Table

The following table contains copy number data on all analyzed exome genes:

  - gene: the gene symbol
  
  - band: the band containing that gene
  
  - del_amp: either deletion or amplification
  
  - num_cases: number of cases with that copy number alteration
  
  - num_samples: number of total samples analyzed
  
  - freq_percent: the alteration frequence in percentage (cases / total samples)
  

```{r plot_table, echo=FALSE}
DT::datatable(
  gene_cn_final_table,
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = TRUE,
)
```
